---
title: "Sustainable Urban Transportation: Exercise #2"
subtitle: "Evaluate Trends in Divvy Bikesharing Trips, 2013-2026"
author: C. Scott Smith, PhD AICP (Instructor)
email: c.scott.smith@depaul.edu
format: 
  html:
    warning: false
    code-tools: true
    code-fold: false
    theme: [flatly, styles.scss]
    toc: true
    toc-depth: 3
    toc-location: left
    reference-location: document
    fontsize: 0.9em
    interlinespace: -0.75em
    embed-resources: false
date: "2026-09-29"
bibliography: references.bib
editor_options:
    chunk_output_type: console
---

```{r}
#| label: activate packages for rendering
#| echo: false

library(dplyr) # data wrangling with pipe syntax
library(tidyverse) # data wrangling
library(sf) # simple features geometry data
library(plotly) # interactive plots
library(leaflet) # interactive maps
library(leaflet.extras) # map reset and full screen buttons
library(data.table) # interactive tables
library(DT) # interactive tables
library(openxlsx) # write tables as Excel spreadsheets

setwd("C:/Users/juste/OneDrive - DePaul University/gitrepos/pedagogy/2027/GEO330_Fall_Quarter/exercises/Exercise_02")

load(file="data/divvy_trips_summary_tables_latest.RData")

```

# Purpose
This project examines trends in Divvy bike sharing ridership over time (i.e., between 2013 and 2026) and by community area within the city of Chicago. Students will explore how ridership varies across communities with respect to demographic and socioeconomic characteristics. The exercise uses custom summary tables produced from a comprehensive dataset of over 56 million Divvy bike trips logged between the official launch of the program on June 27, 2013 through August 31, 2026.

::: {.callout-note collapse="false"}
## Optional readings
National Association of City Transportation Officials (NACTO) [Half a Billion Trips on Shared Micromobility Since 2010](https://nacto.org/publication/shared-micromobility-report-2020-2021/) report [@nacto_halfbillion], which summarizes changes in municipal bikeshare systems over the past decade along with emerging trends in micromobility programs.

Also, consider reading [Dimensions of Divvy: Exploring the performance of bikesharing in Chicago in a period of growth and expansion](https://las.depaul.edu/centers-and-institutes/chaddick-institute-for-metropolitan-development/research-and-publications/Documents/ChaddickInstitute_DivvyReport_Feb2018.pdf) [@chaddick_dimensionsofdivvy]. This study, published in 2018, characterizes the three initial phases of Chicago’s Divvy system beginning with its initial rollout in 2013 through its first and second expansions, in 2015 and 2016, respectively, paying special attention to service and performance gaps.
:::

# Step 1. Open Exercise #2 in Posit Cloud course workspace
Log into the course Posit Cloud “Sustainable Urban Transportation" workspace using the invitation link on the week #2 announcement. Open the Exercise #2 workspace by clicking the corresponding link. A cloud R session will deploy and launch in your web browser.

## Open R script and activate R packages
In your default exercise #2 project you will find an R script named *Exercise_02.R*. You will use this script to input, manage and execute all code for this exercise. The pre-loaded code below is used to activate custom packages that are needed to complete  exercise #2. Activate the R packages by selecting all lines of code beginning with "library" and clicking the Run button located above the top left source/script editor panel.

```{r}
#| label: activate libraries for exercise
#| eval: false

# Activate R packages and load data
library(dplyr) # data wrangling with pipe syntax
library(tidyverse) # data wrangling
library(sf) # simple features geometry data
library(plotly) # interactive plots
library(leaflet) # interactive maps
library(leaflet.extras) # map reset and full screen buttons
library(DT) # interactive tables
library(openxlsx) # write tables as Excel spreadsheets

# Set default options for numerical display
options(scipen=999, digits = 2)

# Load project data
load("data/Exercise_02.RData")

```

<!-- ::: {.callout-tip collapse="true"} -->
<!-- ## VIDEO: Step #1 -->
<!-- {{< video videos/exercise_02_01.mp4 >}} -->
<!-- ::: -->

# Step 2. Download presentation template
*As with previous exercises, you are asked to combine and submit all materials into a single presentation. The basic template for this exercise can be downloaded [via this link to the PowerPoint slides](https://justenvirons.org/pedagogy/GEO330/Exercise_02/slides/Exercise02_Template.pptx).*

This exercise makes use of data from a variety of sources. Custom summary tables of Divvy trip origins and destinations by date and community area between 2013 and 2026 were adapted from raw trip information made available via the City of Chicago's [data portal](https://data.cityofchicago.org/Transportation/Divvy-Trips/fg6s-gzvg) [@chicago_divvytrips] and [Divvy bikeshare system data](https://divvybikes.com/system-data) [@divvy_systemdata] web pages. Frequencies of trips will be mapped by [city of Chicago community area](https://data.cityofchicago.org/Facilities-Geographic-Boundaries/Boundaries-Community-Areas-current-/cauq-8yn6) boundaries [@chicago_communityareas] imported directly from the city of Chicago's data portal. Lastly, variations in trips will be evaluated with respect to neighborhood demographic socioeconomic measures downloaded from the [Chicago Health Atlas](https://chicagohealthatlas.org/download) [@chicago_healthatlas]. *Copy and paste the code below to import the necessary data from the course GitHub and the city of Chicago's data portal.*

<!-- ::: {.callout-tip collapse="true"} -->
<!-- ## VIDEO: Step #2 -->
<!-- {{< video videos/exercise_02_02.mp4 >}} -->
<!-- ::: -->

# Step 3. Explore systemwide trends in Divvy trips over time
Over the past decade the Divvy system has logged over 56 million trips. @fig-cumulativetrips shows the daily cumulative trips from Divvy's launch in June 2013 through August 2026. Dashed vertical lines mark the dates on which the system reached 10, 20, 30, 40 and 50 million trips.

*In slide 1 of your presentation, copy and paste the plot below and describe the cumulative growth of Divvy's systemwide ridership over the past 14 years. How many trips have been logged on Divvy system since its initial rollout? On what date did the system surpass 10 million trips? How long did it take for the Divvy system to log 20, 30, 40 and 50 million trips?*

```{r}
#| label: plot cumulative trips by date
#| echo: true
#| code-fold: true
#| code-summary: "Show the cumulative plotting function"

# function for plotting cumulative trips by date, with vertical lines marking the
# dates the system reached each milestone (10, 20, 30, 40 and 50 million trips)
fx_plot_cumulative_trips <- function(trips_by_date, milestones = seq(10e6, 50e6, by = 10e6)) {

  # first date on which cumulative trips reached each milestone; milestones not
  # yet reached are left out
  trips_by_date <- arrange(trips_by_date, date)
  milestone_rows <- sapply(milestones, function(m) match(TRUE, trips_by_date$trips_cumul >= m))
  milestone_dates <- tibble(trips = milestones,
                            date = trips_by_date$date[milestone_rows]) %>%
    drop_na(date)

  # dashed vertical line for each milestone
  milestone_lines <- lapply(milestone_dates$date, function(d)
    list(type = "line", xref = "x", yref = "paper",
         x0 = d, x1 = d, y0 = 0, y1 = 1,
         line = list(color = "grey", width = 1, dash = "dash")))

  plot_ly(data = trips_by_date,
          x = ~date,
          y = ~trips_cumul,
          name = 'Cumulative trips',
          type = 'scatter',
          mode = 'lines',
          line = list(color = 'lightblue'),
          stackgroup = 'one',
          hovertemplate = 'Cumulative trips: %{y:,}<extra></extra>') %>%
    config(displaylogo = FALSE,
           modeBarButtons = list(list("toImage", "resetScale2d")),
           toImageButtonOptions = list(filename = "divvy_cumulative_trips_by_date")) %>%
    layout(
      shapes = milestone_lines,
      xaxis = list(title = ""),
      yaxis = list(title = "Cumulative Divvy Trips"),
      legend = list(
        font = list(size = 10),
        orientation = "h",
        xanchor = "center",
        x = 0.5, y = -0.1),
      hovermode = "x unified")
}
```

```{r}
#| label: fig-cumulativetrips
#| echo: false
#| fig-cap: Cumulative Divvy Trips, June 2013 through August 2026

# plot cumulative trips by date
fx_plot_cumulative_trips(trips_by_date)
```

Now examine monthly patterns of Divvy ridership over the same period (@fig-tripsbyyearmonth). Winter months (December, January and February) are shown in a darker blue. Hover over each bar to see how ridership differs from month to month. *In slide 2 of your presentation, copy the below figure and note any patterns and/or variations you see in monthly ridership. What factors do you feel explain any differences in ridership by month? What other non-seasonal factors do you feel may have influenced ridership?  Support your findings by referencing statistics from the figure.*

```{r}
#| label: plot trips by year and month
#| echo: true
#| code-fold: true
#| code-summary: "Show the trips by year and month plotting function"

# function for plotting trips by year and month, with winter months (December,
# January and February) in a darker blue than other months
fx_plot_trips_by_yearmonth <- function(trips_by_yearmonth) {
  trips_by_yearmonth <- trips_by_yearmonth %>%
    # lubridate:: keeps data.table's month() from being used when data.table is loaded
    mutate(season = if_else(lubridate::month(yearmonth) %in% c(12, 1, 2), "Winter", "Other"))

  plot_ly() %>%
    add_trace(data = filter(trips_by_yearmonth, season == "Other"),
              x = ~yearmonth,
              y = ~trips,
              type = 'bar',
              name = 'Monthly trips',
              marker = list(color = 'lightblue'),
              hovertemplate = 'Monthly trips: %{y:,}<extra></extra>') %>%
    add_trace(data = filter(trips_by_yearmonth, season == "Winter"),
              x = ~yearmonth,
              y = ~trips,
              type = 'bar',
              name = 'Monthly trips, winter (December to February)',
              marker = list(color = '#08519c'),
              hovertemplate = 'Monthly trips (winter): %{y:,}<extra></extra>') %>%
    config(displaylogo = FALSE,
           modeBarButtons = list(list("toImage", "resetScale2d")),
           toImageButtonOptions = list(filename = "divvy_trips_by_yearmonth")) %>%
    layout(
      barmode = "stack", # winter and other months never share a bar, so bars keep their full width
      xaxis = list(title = ""),
      yaxis = list(title = "Monthly Divvy Trips"),
      legend = list(
        font = list(size = 10),
        orientation = "h",
        xanchor = "center",
        x = 0.5, y = -0.1),
      hovermode = "x unified")
}
```

```{r}
#| label: fig-tripsbyyearmonth
#| echo: false
#| fig-cap: Divvy Trips by Year and Month, June 2013 through August 2026

# plot trips by year and month
fx_plot_trips_by_yearmonth(trips_by_yearmonth)
```

Lastly, let's examine daily patterns of Divvy ridership over the past decade (@fig-tripsbydate). Each point is one day, with weekend days (Saturday and Sunday) shown in a darker blue than weekdays. *In slide 3 of your presentation, copy the below figure and note any patterns and/or variations you see in daily ridership. What factors do you feel explain any daily differences in ridership?*

```{r}
#| label: plot trips by date
#| echo: true
#| code-fold: true
#| code-summary: "Show the trips by date plotting function"

# function for plotting trips by date; each day is a point, with weekend days
# (Saturday and Sunday) in a darker blue than weekdays
fx_plot_trips_by_date <- function(trips_by_date) {
  trips_by_date <- trips_by_date %>%
    # lubridate:: keeps data.table's wday() from being used when data.table is loaded
    mutate(day_type = if_else(lubridate::wday(date) %in% c(1, 7), "Weekend", "Weekday"), # Sunday = 1, Saturday = 7
           day_name = as.character(lubridate::wday(date, label = TRUE, abbr = FALSE)))

  plot_ly() %>%
    add_trace(data = filter(trips_by_date, day_type == "Weekday"),
              x = ~date,
              y = ~trips,
              type = 'scatter',
              mode = 'markers',
              marker = list(color = "lightblue", size = 3),
              name = 'Weekday trips',
              text = ~day_name,
              hovertemplate = '%{text} trips: %{y:,}<extra></extra>') %>%
    add_trace(data = filter(trips_by_date, day_type == "Weekend"),
              x = ~date,
              y = ~trips,
              type = 'scatter',
              mode = 'markers',
              marker = list(color = "#08519c", size = 3),
              name = 'Weekend trips (Saturday and Sunday)',
              text = ~day_name,
              hovertemplate = '%{text} trips: %{y:,}<extra></extra>') %>%
    config(displaylogo = FALSE,
           modeBarButtons = list(list("toImage", "resetScale2d")),
           toImageButtonOptions = list(filename = "divvy_trips_by_date")) %>%
    layout(
      xaxis = list(title = ""),
      yaxis = list(title = "Daily Divvy Trips"),
      legend = list(
        font = list(size = 10),
        orientation = "h",
        xanchor = "center",
        x = 0.5, y = -0.1),
      hovermode = "x unified")
}
```

```{r}
#| label: fig-tripsbydate
#| echo: false
#| fig-cap: Divvy Trips by Day, June 2013 through August 2026

# plot trips by day
fx_plot_trips_by_date(trips_by_date)
```

Finally, examine how long Divvy trips have lasted over time (@fig-tripdurationbyyearmonth). Each box summarizes the daily average trip durations (in minutes per trip) in one month, so each box is based on one observation per day: about 30 observations in a full month (June 2013 has only four, since Divvy launched on June 27). The box spans the middle half of those daily averages, from the 25th percentile (lower edge) to the 75th percentile (upper edge), so it holds about 15 of a month's days. This range is called the interquartile range (IQR). The line inside the box marks the median: half of the month's days had a shorter average trip and half a longer one. In this figure the whiskers extend to the shortest and longest daily averages in each month, and the red line shows each month's overall average (all minutes divided by all trips). Hover over a box to see its values. *In slide 3 of your presentation, note any trends or seasonal patterns in average trip duration. Have trips become shorter or longer over time? What factors (e.g., e-bikes, pricing, membership, weather) might explain the changes?*

```{r}
#| label: plot average trip duration by year and month
#| echo: true
#| code-fold: true
#| code-summary: "Show the trip duration plotting function"

# function for plotting average trip duration by year and month: a box of daily
# averages for each month and a line of each month's overall average
fx_plot_duration_by_yearmonth <- function(duration_by_date, duration_by_yearmonth) {
  plot_ly() %>%
    add_trace(data = duration_by_date %>%
                mutate(yearmonth = floor_date(date, "month")),
              x = ~yearmonth,
              y = ~mins_per_trip,
              type = 'box',
              boxpoints = FALSE, # whiskers extend to min and max
              name = 'Daily average',
              fillcolor = 'lightblue',
              line = list(color = 'steelblue', width = 1)) %>%
    add_trace(data = duration_by_yearmonth,
              x = ~yearmonth,
              y = ~mins_per_trip,
              type = 'scatter',
              mode = 'lines',
              name = 'Monthly average',
              line = list(color = 'darkred', width = 1.5),
              hovertemplate = 'Monthly average: %{y:.1f} min<extra></extra>') %>%
    config(displaylogo = FALSE,
           modeBarButtons = list(list("toImage", "resetScale2d")),
           toImageButtonOptions = list(filename = "divvy_trip_duration_by_yearmonth")) %>%
    layout(
      xaxis = list(title = ""),
      yaxis = list(title = "Average Trip Duration (Minutes per Trip)"),
      legend = list(
        font = list(size = 10),
        orientation = "h",
        xanchor = "center",
        x = 0.5, y = -0.1))
}
```

```{r}
#| label: fig-tripdurationbyyearmonth
#| echo: false
#| fig-cap: Average Divvy Trip Duration (Minutes per Trip) by Year and Month, June 2013 through August 2026

# plot average trip duration by year and month
fx_plot_duration_by_yearmonth(duration_by_date, duration_by_yearmonth)
```

<!-- ::: {.callout-tip collapse="true"} -->
<!-- ## VIDEO: Create custom map -->
<!-- {{< video videos/exercise_02_03.mp4 >}} -->
<!-- ::: -->

# Step 4. Explore total and average trips by community area
In this step we examine variations in Divvy trip patterns across the city by community area (N=77). @fig-frtripsbycommunityarea below shows the distribution of total Divvy trips between 2013 and 2026 across the city. Hover over the community area boundaries to observe their respective total (cumulative) and average daily riderships over the past decade. Daily ridership per 10,000 residents (i.e., ridership standardized by the community area population) is also indicated.

*In slide 4 of your presentation, note any patterns and/or variations you see in total ridership by community area. What factors do you feel explain differences in total and daily ridership? Support your findings by referencing statistics from the figure.*

The R functions below carry out the steps behind @fig-frtripsbycommunityarea: they summarize total trips and average trip durations by community area and map them. The map in Step 5 uses the same functions. You do not need to read or edit this code to complete the exercise, but it is left visible, rather than hidden, so you can see exactly how the map is built. Click "Show the community area trip mapping functions" below to expand it, then copy and run it in your Exercise_02.R script.

```{r}
#| label: define community area trip mapping functions
#| echo: true
#| code-fold: true
#| code-summary: "Show the community area trip mapping functions"

# function for summarizing total (cumulative) trips and average trip duration
# (minutes per trip) by community area over the study period
fx_summarise_communityarea_trips <- function(community_areas_ses, trips_by_yearmonth_ca, duration_by_yearmonth_ca) {

  # cumulative trips as of the latest month
  trips_by_communityarea <- trips_by_yearmonth_ca %>%
    drop_na(yearmonth) %>%
    filter(yearmonth == max(yearmonth))

  # average trip duration (minutes per trip) over the study period
  duration_by_communityarea <- duration_by_yearmonth_ca %>%
    group_by(comarea_id) %>%
    summarise(fr_mins_per_trip = sum(fr_trip_duration_mins, na.rm = TRUE)/sum(fr_trips),
              to_mins_per_trip = sum(to_trip_duration_mins, na.rm = TRUE)/sum(to_trips))

  # join trips and durations to community areas
  community_areas_ses %>%
    left_join(trips_by_communityarea, by = "comarea_id") %>%
    left_join(duration_by_communityarea, by = "comarea_id") %>%
    select(-starts_with("yearmonth"))
}

# function for creating community area hover labels; the selected indicator's
# value and quintile category are added when ses_name is given (Step 5)
fx_communityarea_labels <- function(trips_by_communityarea_ses, total_days, ses_name = NULL) {
  x <- trips_by_communityarea_ses
  labels <- sprintf(
    "<strong>%s</strong><br/>
    <i>Trips (from):</i> %s<br/>
    <i>Trips (to):</i> %s<br/>
    <i>Avg daily trips (from):</i> %0.1f<br/>
    <i>Avg daily trips (to):</i> %0.1f<br/>
    <i>Avg daily trips (from) per 1K pop:</i> %0.1f<br/>
    <i>Avg daily trips (to) per 1K pop:</i> %0.1f<br/>
    <i>Avg trip duration (from):</i> %0.1f min<br/>
    <i>Avg trip duration (to):</i> %0.1f min<br/>",
    x$comarea_name,
    prettyNum(x$fr_trips_cumul, big.mark = ","),
    prettyNum(x$to_trips_cumul, big.mark = ","),
    x$fr_trips_cumul/total_days,
    x$to_trips_cumul/total_days,
    x$fr_trips_cumul/total_days/x$population*1000,
    x$to_trips_cumul/total_days/x$population*1000,
    x$fr_mins_per_trip,
    x$to_mins_per_trip)
  if (!is.null(ses_name)) {
    labels <- paste0(labels,
                     sprintf("<i>%s:</i> %s (%s)<br/>",
                             ses_name,
                             formatC(x$ses_value, format = "f", digits = 1, big.mark = ","),
                             x$ses_ntile))
  }
  lapply(labels, htmltools::HTML)
}

# function for creating the base map shared by the community area maps: basemap,
# reset and full screen buttons and community area boundaries for reference
# (added first, so they are drawn beneath the other layers)
fx_basemap <- function(map_data, carto_api_key = CARTO_API_KEY) {
  leaflet(data = map_data) %>%
    addTiles(urlTemplate = paste0(
      "https://basemaps.cartocdn.com/rastertiles/light_all/{z}/{x}/{y}.png?key=",
      carto_api_key
    )) %>%
    addResetMapButton() %>%
    addFullscreenControl() %>%
    addPolygons(
      fill = FALSE,
      color = "#555",
      weight = 1,
      opacity = 1)
}

# function for mapping total trips (from) by community area (quintiles)
fx_map_trips_by_communityarea <- function(trips_by_communityarea_ses, total_days) {

  # color palette
  map_pal <- colorQuantile(palette = "RdYlBu",
                           domain = trips_by_communityarea_ses$fr_trips_cumul,
                           n = 5,
                           reverse = TRUE)

  # interactive map
  fx_basemap(trips_by_communityarea_ses) %>%
    addPolygons(
      fillColor = ~map_pal(fr_trips_cumul),
      label = fx_communityarea_labels(trips_by_communityarea_ses, total_days),
      weight = 2,
      opacity = 1,
      color = "white",
      dashArray = "3",
      fillOpacity = 0.7) %>%
    addLegend("bottomleft",
              pal = map_pal,
              values = ~fr_trips_cumul,
              title = "Total Divvy Trips (from)<br>Quintiles",
              opacity = 1)
}
```

```{r}
#| label: fig-frtripsbycommunityarea
#| fig-cap: Total Divvy Trips (from) by Community Area (Quintile Categorization), June 2013 through August 2026
#| echo: false

# summarize trips by community area and map total trips
trips_by_communityarea_ses <- fx_summarise_communityarea_trips(community_areas_ses,
                                                               trips_by_yearmonth_ca,
                                                               duration_by_yearmonth_ca)
fx_map_trips_by_communityarea(trips_by_communityarea_ses, total_days = nrow(trips_by_date))
```

<!-- ::: {.callout-tip collapse="true"} -->
<!-- ## VIDEO: Create custom map -->
<!-- {{< video videos/exercise_02_04.mp4 >}} -->
<!-- ::: -->

# Step 5. Explore trends by socioeconomic status
In this step we examine trends in Divvy usage across communities with different socioeconomic characteristics. For this step you will need to modify the below R code to create a custom map, plot and table to inform your analysis. An example analysis is summarized here. 

## Identify and map a community indicator
First, identify one of the seven community indicators downloaded from the [Chicago Health Atlas](https://chicagohealthatlas.org) [@chicago_healthatlas] that you would like to use to explore the equity dimensions of Divvy ridership. Each indicator is stored as a column in the `community_areas_ses` table, which is included in the Divvy summary tables file you loaded in Step 2.

| Column | Indicator | Health Atlas topic | Period |
|:-------|:----------|:-------------------|:-------|
| `population` | Total population | [POP](https://chicagohealthatlas.org/indicators/POP) | 2020-2024 |
| `SVI` | Social Vulnerability Index | [SVI](https://chicagohealthatlas.org/indicators/SVI) | 2022 |
| `poverty` | Poverty rate | [POV](https://chicagohealthatlas.org/indicators/POV) | 2020-2024 |
| `inactivity` | Adult physical inactivity rate | [HCSPAP](https://chicagohealthatlas.org/indicators/HCSPAP) | 2016-2018 |
| `mortality` | All-cause mortality rate | [VRDTHR](https://chicagohealthatlas.org/indicators/VRDTHR) | 2020-2024 |
| `transit` | Ease of walking to transit stop rate | [HCSWTSP](https://chicagohealthatlas.org/indicators/HCSWTSP) | 2023-2024 |
| `violence` | Perceived neighborhood violence rate | [HCSNVP](https://chicagohealthatlas.org/indicators/HCSNVP) | 2023-2024 |

::: {.callout-note collapse="false"}
## Chicago Health Atlas
The [Chicago Health Atlas](https://chicagohealthatlas.org) publishes data on hundreds of sociodemographic and public health indicators for communities throughout the city. The website  allows for searching and exploring various metrics via maps, charts and tables. Indicator data can also be downloaded by community area, census tract and zip code. Refer to the Atlas' [how to use](https://chicagohealthatlas.org/how-to) page for more information.  
:::

Follow the topic links for a full description of each indicator, including its units and data source. Community areas are grouped into five quintile categories (Lowest, Low, Moderate, High and Highest) of the indicator's values. Before interpreting your results, note whether higher values of your indicator signal more or less advantaged conditions (e.g., a higher poverty rate or social vulnerability index signals less advantaged conditions).

The R functions below carry out the steps behind @fig-mapsesquintile: they add your indicator's values and quintile categories to the community areas and map them along with Divvy stations. They use the community area functions from Step 4, so run those first. You do not need to read or edit this code to complete the exercise, but it is left visible, rather than hidden, so you can see exactly how the map is built. Click "Show the indicator mapping functions" below to expand it, then copy and run it in your Exercise_02.R script.

*After selecting one of the indicators, copy and paste the following code into your R script. Edit the two lines marked `EDIT HERE` with your indicator's name and column (from the table above) and follow the instructions in the video to create a custom map. The map also shows current Divvy stations [@divvy_gbfs], grouped into numbered clusters when zoomed out. Use the layer control in the top right corner to turn the stations and community areas on and off. Copy the map to slide 5 of your presentation and describe any visual relationship you see (or don't see) between ridership, station locations and the indicator you selected.*


```{r}
#| label: define indicator mapping functions
#| echo: true
#| code-fold: true
#| code-summary: "Show the indicator mapping functions"

# function for adding the selected indicator's values (ses_value) and quintile
# categories (ses_ntile) to community areas
fx_add_ses_quintiles <- function(community_areas_ses, ses_column) {
  quintile_labels <- c("Lowest", "Low", "Moderate", "High", "Highest")
  community_areas_ses %>%
    mutate(ses_value = .data[[ses_column]],
           ses_ntile = quintile_labels[ntile(ses_value, 5)])
}

# function for mapping the selected indicator by community area (quintiles) with
# Divvy stations; a layer control turns the community areas and stations on and off
fx_map_ses_quintiles <- function(trips_by_communityarea_ses, divvy_stations, ses_name, total_days) {

  # color palette
  map_pal <- colorQuantile(palette = "RdYlBu",
                           domain = trips_by_communityarea_ses$ses_value,
                           n = 5,
                           reverse = TRUE)

  # station hover labels
  station_labels <- sprintf(
    "<strong>%s</strong><br/>
    <i>Docks:</i> %s<br/>
    <i>Community area:</i> %s",
    divvy_stations$name,
    divvy_stations$capacity,
    coalesce(divvy_stations$comarea_name, "Outside Chicago")) %>%
    lapply(htmltools::HTML)

  # interactive map
  fx_basemap(trips_by_communityarea_ses) %>%
    addPolygons(
      fillColor = ~map_pal(ses_value),
      label = fx_communityarea_labels(trips_by_communityarea_ses, total_days, ses_name),
      weight = 2,
      opacity = 1,
      color = "white",
      dashArray = "3",
      fillOpacity = 0.7,
      group = "Community Areas") %>%
    # stations are grouped into numbered clusters when zoomed out
    addCircleMarkers(
      data = divvy_stations,
      radius = 5,
      color = "#fff",
      weight = 1,
      fillColor = "steelblue",
      fillOpacity = 0.9,
      label = station_labels,
      # clusters drawn as steelblue circles showing the number of stations,
      # replacing the default green, yellow and orange size classes
      clusterOptions = markerClusterOptions(
        showCoverageOnHover = FALSE, # no outline of the cluster's area on hover
        iconCreateFunction = htmlwidgets::JS("
          function(cluster) {
            return new L.DivIcon({
              html: '<div style=\"background-color: rgba(70, 130, 180, 0.85); color: white; font-weight: bold; border: 2px solid white; border-radius: 50%; width: 36px; height: 36px; line-height: 32px; text-align: center;\">' + cluster.getChildCount() + '</div>',
              className: '',
              iconSize: new L.Point(36, 36)
            });
          }")),
      group = "Divvy Stations") %>%
    addLegend("bottomleft",
              pal = map_pal,
              values = ~ses_value,
              title = paste0(ses_name, "<br>Quintiles"),
              opacity = 1,
              group = "Community Areas") %>%
    addLayersControl(
      overlayGroups = c("Community Areas", "Divvy Stations"),
      options = layersControlOptions(collapsed = FALSE))
}
```

```{r}
#| label: fig-mapsesquintile
#| fig-cap: Selected Community Indicator by Community Area (Quintile Categorization) with Divvy Stations
#| echo: true

# Edit the following two lines with the selected indicator's name and column
ses_name <- "Poverty" # EDIT HERE: indicator name used in map labels and legend
ses_column <- "poverty" # EDIT HERE: population, SVI, poverty, inactivity, mortality, transit or violence

# add the selected indicator's values and quintile categories to community areas
community_areas_ses <- fx_add_ses_quintiles(community_areas_ses, ses_column)

# summarize trips by community area and map the indicator with Divvy stations
trips_by_communityarea_ses <- fx_summarise_communityarea_trips(community_areas_ses,
                                                               trips_by_yearmonth_ca,
                                                               duration_by_yearmonth_ca)
fx_map_ses_quintiles(trips_by_communityarea_ses, divvy_stations, ses_name,
                     total_days = nrow(trips_by_date))
```

## Examine Divvy ridership trends by socioeconomic indicator over time
Now you will evaluate Divvy ridership over the past decade for communities with different socioeconomic characteristics. The R functions below carry out the steps behind @fig-tripsbyses: they summarize trips by year, month and quintile category of your indicator and plot them. You do not need to read or edit this code to complete the exercise, but it is left visible, rather than hidden, so you can see exactly how the plot is built. Click "Show the indicator trend functions" below to expand it, then copy and run it in your Exercise_02.R script.

*Using the same indicator you selected above, copy and paste the following code chunk into your R script and follow the instructions in the video to create a custom plot. Copy the plot into slide 6 of your presentation and describe any relationships you see (or don't see) between ridership and the socioeconomic indicator you selected over the past ten years. Have variations in ridership changed differently over time with respect to community socioeconomic condition?*

```{r}
#| label: define indicator trend functions
#| echo: true
#| code-fold: true
#| code-summary: "Show the indicator trend functions"

# quintile category colors, from Highest to Lowest (also used in Step 6)
quintile_colors <- c(Highest = "darkred",
                     High = "orange",
                     Moderate = "yellow",
                     Low = "lightblue",
                     Lowest = "darkblue")

# function for summarizing trips by year, month and quintile category of the
# selected indicator (added to community areas with fx_add_ses_quintiles)
fx_summarise_trips_by_ses <- function(trips_by_yearmonth_ca, community_areas_ses, total_days) {
  trips_by_yearmonth_ca %>%
    left_join(community_areas_ses %>% st_drop_geometry(), by = "comarea_id") %>%
    group_by(yearmonth, ses_ntile) %>%
    summarise(tot_pop = sum(population),
              fr_trips = sum(fr_trips),
              to_trips = sum(to_trips),
              fr_trips_cumul = sum(fr_trips_cumul),
              to_trips_cumul = sum(to_trips_cumul),
              fr_dailytrips = fr_trips/total_days,
              to_dailytrips = to_trips/total_days,
              fr_trips_per1k = fr_trips/tot_pop*1000,
              to_trips_per1k = to_trips/tot_pop*1000,
              .groups = "drop") %>%
    drop_na(tot_pop) %>%
    mutate(order = case_when(ses_ntile == "Lowest" ~ 1,
                             ses_ntile == "Low" ~ 2,
                             ses_ntile == "Moderate" ~ 3,
                             ses_ntile == "High" ~ 4,
                             ses_ntile == "Highest" ~ 5),
           ses_label = factor(ses_ntile, levels = names(quintile_colors))) %>%
    arrange(yearmonth, order)
}

# function for plotting monthly trips per 1,000 residents by quintile category;
# ses_name labels the legend
fx_plot_trips_by_ses <- function(trips_by_yearmonth_by_ses, ses_name = "Indicator") {
  plot_ly(data = trips_by_yearmonth_by_ses,
          x = ~yearmonth,
          y = ~fr_trips_per1k,
          type = 'scatter',
          mode = 'lines',
          color = ~ses_label,
          colors = unname(quintile_colors),
          text = ~ses_label,
          hovertemplate = '%{text}: %{y:0.1f}<extra></extra>') %>%
    config(displaylogo = FALSE,
           modeBarButtons = list(list("toImage", "resetScale2d")),
           toImageButtonOptions = list(filename = paste0("divvy_trips_by_", tolower(gsub("[^A-Za-z0-9]+", "_", ses_name)), "_quintile"))) %>%
    layout(
      xaxis = list(title = "Year and Month"),
      yaxis = list(title = "Monthly Divvy Trips (from) per 1,000 Residents"),
      legend = list(
        title = list(text = paste0(ses_name, " Quintile  ")),
        font = list(size = 10),
        xanchor = "center",
        orientation = "h",
        x = 0.5, y = -0.1),
      hovermode = "x unified")
}
```

```{r}
#| label: fig-tripsbyses
#| fig-cap: Monthly Divvy Trips per 1,000 Residents by Quintile Category of the Selected Community Indicator
#| echo: true

# summarize trips by quintile category of the selected indicator and plot them
trips_by_yearmonth_by_ses <- fx_summarise_trips_by_ses(trips_by_yearmonth_ca,
                                                       community_areas_ses,
                                                       total_days = nrow(trips_by_date))
fx_plot_trips_by_ses(trips_by_yearmonth_by_ses, ses_name)
```

## Summary of ridership by socioeconomic indicator 
Here you create a final custom table with your summary information by socioeconomic category. The R functions below summarize trips and population by quintile category and format the table (@tbl-tripsbysestable). You do not need to read or edit this code to complete the exercise, but it is left visible, rather than hidden, so you can see exactly how the table is built. Click "Show the summary table functions" below to expand it, then copy and run it in your Exercise_02.R script.

*Copy, paste and run the following code chunks in your R script. The second chunk saves the table to your project data directory for further analysis. Copy and paste the table into slide 6 of your presentation (along with the above plot) and briefly summarize your findings.*

```{r}
#| label: define summary table functions
#| echo: true
#| code-fold: true
#| code-summary: "Show the summary table functions"

# function for summarizing population and trips by quintile category of the
# selected indicator, from Highest to Lowest
fx_summarise_ses_quintiles <- function(trips_by_yearmonth_by_ses) {
  trips_by_yearmonth_by_ses %>%
    group_by(ses_ntile) %>%
    summarise(order = max(order),
              tot_pop = max(tot_pop),
              fr_trips = sum(fr_trips),
              to_trips = sum(to_trips)) %>%
    mutate(tot_pop_pct = tot_pop/sum(tot_pop),
           fr_trips_pct = fr_trips/sum(fr_trips),
           to_trips_pct = to_trips/sum(to_trips)) %>%
    arrange(-order) %>%
    select(-order)
}

# function for formatting the summary table as an interactive table; ses_name
# labels the quintile column
fx_table_ses_quintiles <- function(trips_by_ses_quintile, ses_name = "Indicator") {
  trips_by_ses_quintile %>%
    datatable(class = 'cell-border stripe',
              rownames = FALSE,
              colnames = c(paste(ses_name, "Quintile"), "Total Population", "Trips (from)", "Trips (to)",
                           "Percent of Population", "Percent of Trips (from)", "Percent of Trips (to)"),
              options = list(pageLength = 5)) %>%
    formatPercentage(c("tot_pop_pct", "fr_trips_pct", "to_trips_pct"), digits = 1) %>%
    formatRound(c("tot_pop", "fr_trips", "to_trips"), mark = ",", digits = 0)
}
```

```{r}
#| label: tbl-tripsbysestable
#| tbl-cap: Divvy Trips (6/27/2013-8/31/2026) and Population by Quintile Category of the Selected Community Indicator
#| echo: true

# summarize trips by quintile category and display the table
trips_by_ses_quintile <- fx_summarise_ses_quintiles(trips_by_yearmonth_by_ses)
fx_table_ses_quintiles(trips_by_ses_quintile, ses_name)
```

```{r}
#| label: save table summarizing trips by socioeconomic category
#| eval: false
#| echo: true

# save the summary table as an Excel spreadsheet in your project data directory
write.xlsx(trips_by_ses_quintile,
           file = "data/trips_by_ses_quintile.xlsx",
           asTable = TRUE,
           overwrite = TRUE)
```

<!-- ::: {.callout-tip collapse="true"} -->
<!-- ## VIDEO: Create custom map -->
<!-- {{< video videos/exercise_02_05.mp4 >}} -->
<!-- ::: -->

# Step 6. Compare community areas across socioeconomic categories
In this final step you will take a closer look at five individual community areas, one from each of the socioeconomic quintile categories you used in Step 5 (i.e., Lowest, Low, Moderate, High and Highest). Use the map from Step 5 (@fig-mapsesquintile) to choose them.

The R functions below look up your selected community areas and create @fig-tripsbyyearmonthselected and @fig-tripdurationselected. They use the quintile categories and colors from Step 5, so run those functions first. You do not need to read or edit this code to complete the exercise, but it is left visible, rather than hidden, so you can see exactly how the figures are built. Click "Show the selected community area functions" below to expand it, then copy and run it in your Exercise_02.R script.

*Copy and paste the following code chunk into your R script. Replace the five community area names with the ones you selected, one from each quintile category, and run the code. Check the table (@tbl-selectedcommunityareas) to make sure each quintile category appears once.*

```{r}
#| label: define selected community area functions
#| echo: true
#| code-fold: true
#| code-summary: "Show the selected community area functions"

# function for looking up the selected community areas, their quintile categories
# and colors (from Highest to Lowest)
fx_select_communityareas <- function(community_areas_ses, selected_communityareas, colors = quintile_colors) {
  community_areas_ses %>%
    st_drop_geometry() %>%
    filter(comarea_name %in% selected_communityareas) %>%
    select(comarea_id, comarea_name, ses_ntile) %>%
    mutate(ses_label = factor(ses_ntile, levels = names(colors)),
           ca_label = paste0(comarea_name, " (", ses_ntile, ")"),
           color = unname(colors[as.character(ses_ntile)])) %>%
    arrange(ses_label)
}

# function for plotting monthly trips (from) for the selected community areas
fx_plot_trips_selected <- function(trips_by_yearmonth_ca, selected_ca) {
  trips_by_yearmonth_selected <- trips_by_yearmonth_ca %>%
    inner_join(selected_ca, by = "comarea_id") %>%
    mutate(ca_label = factor(ca_label, levels = selected_ca$ca_label))

  plot_ly(data = trips_by_yearmonth_selected,
          x = ~yearmonth,
          y = ~fr_trips,
          type = 'scatter',
          mode = 'lines',
          color = ~ca_label,
          colors = setNames(selected_ca$color, selected_ca$ca_label),
          text = ~ca_label,
          hovertemplate = '%{text}: %{y:,}<extra></extra>') %>%
    config(displaylogo = FALSE,
           modeBarButtons = list(list("toImage", "resetScale2d")),
           toImageButtonOptions = list(filename = "divvy_trips_by_yearmonth_selected_communityareas")) %>%
    layout(
      xaxis = list(title = "Year and Month"),
      yaxis = list(title = "Monthly Divvy Trips (from)"),
      legend = list(
        title = list(text = "Community Area (Quintile)  "),
        font = list(size = 10),
        orientation = "h",
        xanchor = "center",
        x = 0.5, y = -0.1),
      hovermode = "x unified")
}

# function for plotting a box plot of monthly average trip duration (from) for
# the selected community areas
fx_plot_duration_selected <- function(duration_by_yearmonth_ca, selected_ca) {
  duration_by_yearmonth_selected <- duration_by_yearmonth_ca %>%
    inner_join(selected_ca, by = "comarea_id") %>%
    mutate(ca_label = factor(ca_label, levels = selected_ca$ca_label))

  plot_ly(data = duration_by_yearmonth_selected,
          x = ~ca_label,
          y = ~fr_mins_per_trip,
          type = 'box',
          boxpoints = "outliers", # whiskers extend to 1.5 x IQR; months beyond them are plotted as points
          color = ~ca_label,
          colors = setNames(selected_ca$color, selected_ca$ca_label)) %>%
    config(displaylogo = FALSE,
           modeBarButtons = list(list("toImage", "resetScale2d")),
           toImageButtonOptions = list(filename = "divvy_trip_duration_selected_communityareas")) %>%
    layout(
      xaxis = list(title = "Community Area (Quintile)"),
      yaxis = list(title = "Average Trip Duration (Minutes per Trip)",
                   hoverformat = ".1f"), # hover values to tenths
      showlegend = FALSE)
}
```

```{r}
#| label: tbl-selectedcommunityareas
#| tbl-cap: Selected Community Areas by Quintile Category
#| echo: true

# EDIT HERE: replace with five community areas, one from each quintile category
selected_communityareas <- c("Near South Side",
                             "Near North Side",
                             "Austin",
                             "Bridgeport",
                             "Englewood")

# look up selected community areas and their quintile categories
selected_ca <- fx_select_communityareas(community_areas_ses, selected_communityareas)
selected_ca %>%
  select(`Community Area` = comarea_name, `Quintile Category` = ses_ntile) %>%
  knitr::kable()
```

## Trips by year and month for selected community areas
*Run the code below to plot monthly trips starting in each of your selected community areas (@fig-tripsbyyearmonthselected). Copy the plot into slide 7 of your presentation and describe how ridership has changed over time in each community area. Do the trends differ across socioeconomic categories?*

```{r}
#| label: fig-tripsbyyearmonthselected
#| fig-cap: Monthly Divvy Trips (from) for Selected Community Areas, June 2013 through August 2026
#| echo: true

# plot monthly trips for the selected community areas
fx_plot_trips_selected(trips_by_yearmonth_ca, selected_ca)
```

## Average trip duration for selected community areas
@fig-tripdurationselected uses the same kind of box plot as @fig-tripdurationbyyearmonth, but here each observation is one month's average trip duration (minutes per trip) for trips starting in a community area. Each box includes every month in which at least one trip started there, up to 159 months (June 2013 through August 2026), so community areas that received Divvy stations after 2013 have fewer months. Unlike the earlier figure, the whiskers extend only to the shortest and longest monthly averages within 1.5 times the interquartile range of the box. Months beyond the whiskers are plotted as individual points called outliers: months whose average trip duration was unusually short or long for that community area, often because only a few trips were taken (e.g., in winter or soon after stations opened).

*Run the code below to create the box plot. Copy the plot into slide 8 of your presentation. How do trip durations compare across the five community areas? Which community areas have the most outliers, and what might explain them? What might explain the differences between community areas?*

```{r}
#| label: fig-tripdurationselected
#| fig-cap: Monthly Average Divvy Trip Duration (Minutes per Trip) for Selected Community Areas, June 2013 through August 2026
#| echo: true

# plot a box plot of average trip duration for the selected community areas
fx_plot_duration_selected(duration_by_yearmonth_ca, selected_ca)
```

<!-- ::: {.callout-tip collapse="true"} -->
<!-- ## VIDEO: Create custom map -->
<!-- {{< video videos/exercise_02_06.mp4 >}} -->
<!-- ::: -->

# References

This exercise draws on several datasets published by the City of Chicago, the Chicago Department of Public Health and Divvy. Each is widely used by transportation planners, public health practitioners and researchers in Chicago, and each is worth understanding on its own terms.

**Divvy trip data.** Divvy is Chicago's bikeshare system, which launched on June 27, 2013. Each Divvy trip record includes the trip's start and end times and locations, along with details such as the rider type (member or casual). Trips from 2013 through 2019 were downloaded from the City of Chicago's Divvy Trips dataset [@chicago_divvytrips], and trips from 2020 onward from the monthly files on Divvy's system data page [@divvy_systemdata]. The more than 56 million trips were then summarized by date and by community area of trip origin and destination to create the summary tables used in this exercise.

**Divvy stations.** Station locations come from Divvy's General Bikeshare Feed Specification (GBFS) feed [@divvy_gbfs]. GBFS is an open data standard, maintained by MobilityData [@mobilitydata_gbfs], that bikeshare and other shared mobility systems use to publish information about their stations and vehicles. The station information feed lists each station's name, location and number of docks. Because the feed describes the system as it is today, the stations shown in Step 5 are current stations, not the stations that existed when earlier trips were taken.

**Chicago community areas.** Trips are mapped and summarized by the city's 77 community areas [@chicago_communityareas], downloaded from the City of Chicago's data portal. Community area boundaries have remained largely unchanged for decades, which makes them a common, stable unit for comparing neighborhoods and tracking change over time.

**Chicago Health Atlas.** The community indicators used in Steps 5 and 6 come from the Chicago Health Atlas [@chicago_healthatlas], which publishes hundreds of sociodemographic and public health indicators by community area, census tract and zip code. Each indicator's page on the Atlas describes its definition, units, data source and time period.

**Basemap tiles.** The interactive maps in this exercise are built on CARTO's Positron basemap tiles [@carto_basemaps], a minimal, light-colored basemap designed to keep thematic data (such as the community area polygons and Divvy stations shown here) legible in the foreground.

This exercise also points to supplementary readings that are not code inputs but provide background on bikesharing in Chicago and beyond.

**Shared micromobility trends.** NACTO's *Half a Billion Trips on Shared Micromobility Since 2010* [@nacto_halfbillion] summarizes the growth of bikeshare and scooter share systems across North America, and is suggested as optional background reading before Step 1.

**Divvy's growth and expansion.** The Chaddick Institute's *Dimensions of Divvy* [@chaddick_dimensionsofdivvy] characterizes Divvy's initial rollout in 2013 and its expansions in 2015 and 2016, paying special attention to service and performance gaps across Chicago's neighborhoods.

::: {#refs}
:::
