---
title: "Sustainable Urban Transportation (GEO 3/430) Exercise #1"
subtitle: "Transportation Mode Share, Pedestrian Fatalities and Urban Street Design"
author: C. Scott Smith, PhD AICP (Instructor)
email: c.scott.smith@depaul.edu
format:
  html:
    css: styles.scss
    warning: false
    code-tools: false
    code-fold: false
    echo: false
    eval: true
    theme: flatly
    toc: true
    toc-depth: 3
    toc-location: left
    reference-location: document
    fontsize: 0.9em
    interlinespace: -0.75em
    embed-resources: true
date: 2026-9-9
bibliography: ../references.bib
editor_options:
  chunk_output_type: console
---

```{r}
#| label: set up environment
#| eval: true
#| echo: false

# Set the chunk working directory to this exercise's own folder (the
# parent of the "scripts" folder this document lives in), computed from
# the document's own location rather than hardcoded. This makes the
# render work the same way on any machine or Posit Cloud project,
# regardless of what folder the render was started from. Students do not
# need this in their own Exercise_01.R script: in the Posit Cloud
# project, the working directory is already the exercise folder.

knitr::opts_knit$set(root.dir = dirname(dirname(knitr::current_input(dir = TRUE))))
```

# Purpose
This exercise uses two nationwide datasets, commuting (or "journey to work") mode share and pedestrian fatality records, to explore transportation mode share trends and safety patterns within a US county that you select. Working between 2010 and 2024, you will build maps, tables and figures that describe how people get to work, where pedestrian fatalities occur, and how these patterns have changed over time. You will then propose practical ideas for improving active transportation (walking, bicycling and transit) and reducing pedestrian injury in your chosen county.

Understanding who walks, bikes and takes transit, and where people face the greatest safety risk while doing so, sits at the center of sustainable urban transportation planning. The figures and maps you create in this exercise are the same types of evidence that planners, engineers and advocates use to justify investments in safer streets, better transit service and more walkable, bikeable neighborhoods.

::: {.callout-note collapse="false"}
## Optional Reading
- [Dangerous by Design](https://www.smartgrowthamerica.org/signature-reports/dangerous-by-design/) [@sga_dangerousbydesign] (Smart Growth America), produced with the National Complete Streets Coalition (a program of Smart Growth America) and with data analysis support from the State Smart Transportation Initiative, ranks all 50 states and the 101 largest U.S. metro areas by pedestrian fatality rate. The report's principal data source is the Fatality Analysis Reporting System (FARS), maintained by the National Highway Traffic Safety Administration (NHTSA). Smart Growth America uses these data, along with population data from the US Census Bureau's American Community Survey (ACS), to determine pedestrian death counts and rates by state and metro area for the report's rankings.
:::

# Step 1. Open Exercise #1 in Posit Cloud course workspace
Log into the course Posit Cloud "Sustainable Urban Transportation" workspace using the invitation link on the week #1 announcement. Open the Exercise #1 workspace by clicking the corresponding link. A cloud R session will launch in your web browser.

::: {.callout-note collapse="false"}
## Concerning coding
This class does not have a coding prerequisite, and no prior programming experience is required. Students will not be asked to write or prepare code from scratch at any point in the course. Instead, coding tools are used simply as a means to create flexible, interactive, and customizable learning activities. This approach allows students to explore different topics and areas of interest, and to integrate and work with a wide variety of datasets in ways that would otherwise be difficult to achieve. The goal is to enhance the learning experience, not to teach programming itself.
:::

## Open R script, activate R packages and load project data
In your default Exercise #1 project you will find an R script named *Exercise_01.R*. You will use this script to enter, manage and run all code for this exercise. The code below, which has already been copied into the Exercise_01.R script, activates the custom packages needed to complete Exercise #1. Run it by selecting all lines that begin with "library" and clicking the Run button located above the top left source or script editor panel. 

*Purpose of the code below: activate the R packages required for the exercise and load the project's prepared dataset into your R session. You will use several R packages to complete the exercise, including "dplyr" and "tidyverse" for data formatting and wrangling, "sf" (simple features) for importing and transforming geographic data, "plotly" for creating interactive figures, and "leaflet" for producing interactive maps.*

```{r}
#| label: activate r packages and load project data
#| echo: true
#| eval: true

# Activate R packages
library(dplyr) # data wrangling with pipe syntax
library(tidyverse) # data wrangling
library(sf) # simple features geometry data
library(plotly) # interactive plots
library(leaflet) # interactive maps
library(leaflet.extras) # additional leaflet controls, such as fullscreen and reset view buttons
library(leafsync) # synchronizes panning and zooming across multiple leaflet maps shown side by side
library(crsuggest) # suggests an appropriate projected coordinate system for a given area
library(spatstat.geom) # defines the spatial point pattern objects used in spatial analysis
library(spatstat.explore) # calculates kernel density estimates and other exploratory spatial statistics
library(terra) # processes raster data, such as the pedestrian fatality density surface
library(classInt) # calculates class intervals (e.g., natural breaks) for shading maps

# Set default options for numerical display
options(scipen=999, digits = 2)

# Load project data
load("data/Exercise_01.RData")

```

::: {.callout-note collapse="false"}
## What is R?
R is a free software environment for statistical computing and graphics. It compiles and runs on a wide variety of UNIX platforms, Windows and MacOS, and is also available through a cloud version, *Posit Cloud*. RStudio is an integrated development environment (IDE) for R. It provides a set of tools built to help R users be more productive with R (and other languages, including Python). It includes a console and a syntax highlighting editor that supports direct code execution, along with tools for plotting, viewing history, debugging and managing your workspace.
:::

<!-- ::: {.callout-tip collapse="true"} -->
<!-- ## VIDEO: Step #1 -->
<!-- {{< video videos/exercise_01_01.mp4 >}} -->
<!-- ::: -->

# Step 2. Download exercise template and select US county
Download the [exercise PowerPoint template](https://justenvirons.org/pedagogy/GEO330/Exercise_01/slides/Exercise01_Template.pptx) using the link provided. You will draw on resources from this page, such as the maps and plots below, to complete your presentation. For now, just download the template and store locally on your machine for editing.

For this exercise, choose a large (population over 500,000), urban county located in the contiguous United States as your study area. Browse the interactive map of US counties below (@fig-popbycountymap) to identify a study area that interests you. Note the county's name and its unique "GEOID". The GEOID combines a two digit state code with a three digit county code. For example, 17031 is the GEOID for Cook County, Illinois. Selecting a county with a large population gives you enough commuters and enough recorded crashes to see meaningful patterns rather than noise from a small sample, which matters whenever you are trying to draw conclusions that could inform real transportation policy.

Population counts shown on the map come from the American Community Survey (ACS) 5 year estimates [@census_acs], the same nationwide survey described in more detail in Step 3. This map, and every other interactive map in this exercise, is drawn on CARTO's Positron basemap tiles [@carto_basemaps], with Esri's World Imagery satellite tiles [@esri_worldimagery] available as an alternate layer.

```{r}
#| label: fig-popbycountymap
#| fig-cap: Total Population by US County
#| eval: true
#| code-fold: true

# Make interactive map of US counties by population using leaflet package

# assign bins for population thematic map loosely based on quintile breaks
counties_bins <- c(0, 25000, 100000, 500000, 999999, Inf)

# create palette using colorblind/accessible colors
counties_pal <- colorBin(
  palette="viridis",
  domain=us_counties_population_geo$total_population,
  na.color="transparent",
  bins = counties_bins,
  reverse = TRUE
  )

# format/generate popup labels for map
labels = sprintf(
  "<strong>%s County, %s</strong><br/>
  Population (2024): %s<br/>
  GEOID: %s",
  us_counties_population_geo$NAME,
  us_counties_population_geo$STATE_NAME,
  format(us_counties_population_geo$total_population, big.mark = ","),
  us_counties_population_geo$GEOID) %>%
  lapply(htmltools::HTML)

# create thematic map using leaflet
leaflet() %>%
  # addProviderTiles(providers$CartoDB.Positron) %>%
    addTiles(urlTemplate = paste0(
      "https://basemaps.cartocdn.com/rastertiles/light_all/{z}/{x}/{y}.png?key=",
      CARTO_API_KEY
    )) %>%
    addProviderTiles(providers$Esri.WorldImagery, group = "World Imagery (satellite)") %>%
  addPolygons(group = "Counties",
              data=us_counties_population_geo %>% st_transform(crs=4326),
              # fillColor = "orange",
              fillColor = ~counties_pal(total_population),
              weight = 0.5,
              opacity = 0.5,
              color = "white",
              dashArray = "3",
              fillOpacity = 0.7,
              highlight = highlightOptions(
                weight = 5,
                color = "#666",
                dashArray = "",
                fillOpacity = 1,
                bringToFront = FALSE),
              label = labels,
              labelOptions = labelOptions(
                style = list("font-weight" = "normal", padding = "3px 8px"),
                textsize = "15px",
                direction = "auto")) %>%
  addPolylines(
      data = us_states_sub_geom,
      stroke = TRUE,
      weight = 2,
      opacity = 1,
      fillOpacity = 0,
      color = "#FFF",
      group = "States"
    ) %>%
    addPolylines(
      data = us_divisions_sub_geom,
      stroke = TRUE,
      weight = 2,
      opacity = 1,
      fillOpacity = 0,
      color = "black",
      group = "Divisions"
    ) %>%
  addLegend("bottomleft", pal = counties_pal, values = us_counties_population_geo$total_pop,
            title = paste0("Population (2024)"),
            opacity = 1) %>%
  addScaleBar(position = "bottomright") %>%
    addFullscreenControl() %>%
  addResetMapButton() %>% 
    addLayersControl(
      baseGroups = c(# "Basemap (default)",
        "Positron (minimal)", "World Imagery (satellite)"),
      overlayGroups = c("Counties", "States", "Divisions"),
      options = layersControlOptions(collapsed = FALSE)
    )

```

<!-- ::: {.callout-tip collapse="true"} -->
<!-- ## VIDEO: Step #2 -->
<!-- {{< video videos/exercise_01_02.mp4 >}} -->
<!-- ::: -->

# Step 3. Transportation mode share analysis
In this step, you will examine trends in commute mode share across US counties over the past fifteen years using a curated dataset. The data were downloaded using the [Census application programming interface (API)](https://www.census.gov/data/developers/data-sets.html) in RStudio (code available upon request). The Census API supports custom queries and downloads of American Community Survey (ACS) and other Census Bureau data.

The commute mode share data for US counties come from table [B08301](https://censusreporter.org/tables/B08301/) [@census_b08301] of the American Community Survey (ACS) 5 year estimates [@census_acs]. Because the ACS is collected continuously, using the same questions and methods nationwide, it is one of the primary sources transportation planners, engineers and researchers use to track commuting behavior, evaluate demand for walking, bicycling and transit investment, and monitor progress toward sustainable transportation goals over time.

The Census Bureau asks a "journey to work" question that supports a variety of planning and policy purposes, including those related to sustainable transportation. The exact wording of the question is "How did this person usually get to work last week?" A respondent who used more than one mode of transportation was [asked to select the mode](https://www2.census.gov/programs-surveys/acs/about/qbyqfact/2016/JourneytoWork.pdf) [@census_journeytowork] used for *most* of the distance traveled. For example, a respondent would select "bicycle" or "walk" only if that mode covered most of the commute distance. Because of this, the ACS captures only each person's primary commuting mode during the single week before the survey, which may not fully represent their travel behavior across the year. Even with this limitation, the ACS remains the most reliable nationwide source for tracking change over time in commute mode share.

Commute mode share is one of the clearest signals of how sustainable a region's transportation system is. A county where walking, bicycling and transit make up a growing share of commutes, for example, is generally reducing its per capita vehicle emissions, traffic congestion and demand for parking.

## Trends in mode share

Examine @fig-modesharetrends below, which shows trends for selected commute modes (from home, walk, transit and bicycle) across the United States between 2010 and 2024. (Note that drive alone is not included in the plot, since its disproportionate share tends to obscure any movement among modes with less representation.) Copy and paste this figure and @fig-modeshareboxplot into slide #1 of your presentation and summarize the trends you observe.

```{r}
#| label: fig-modesharetrends
#| fig-cap: Average Commute Mode Share among US Counties by Select Modes, 2010 to 2024
#| fig-subcap:
#|   - "Mode Share by Select Modes"
#|   - "Total Walk Commuters by Year"
#| layout-ncol: 2
#| eval: true
#| echo: false

fx_plot_modeshare("all") # all counties combined

fx_plot_activetrans("all") # total walk commuters by year, all counties combined

```

*Now copy and paste the following code at the end of your own Exercise_01.R script. Execute the script to build the same kind of plot for the county you selected county. @fig-modesharetrendscook shows such a plot for Cook County, Illinois, as a worked example. Essentially replace the GEOID value with the code for your selected county to generate your own version of the figure. Copy and paste your county's version of the figure into slide #2 of your presentation, summarize the trends you see, and compare them with the nationwide pattern shown in @fig-modesharetrends and the other figures below.*

::: {.callout-note collapse="false"}
## Copying and pasting code
Each time you work through your script, be sure to copy and paste new code *at the end of the file*. This step is important for ensuring your progress is saved and executed correctly. Even though you are not writing code from scratch, copying and pasting each successive code chunk at the end of your Exercise #1 script is critical as skipping this step may result in errors, incomplete or missing work.
:::

```{r}
#| label: fig-modesharetrendscook
#| fig-cap: Average Commute Mode Share for Cook County, Illinois, 2010 to 2024
#| fig-subcap:
#|   - "Mode Share by Select Modes"
#|   - "Total Walk Commuters by Year"
#| layout-ncol: 2
#| echo: true

custom_geoid <- "17031" # EDIT HERE! Replace with the GEOID for the county you selected

# run the following function to create a custom transportation mode share plot
fx_plot_modeshare(custom_geoid)

fx_plot_activetrans(custom_geoid) # total walk commuters by year
```

## Mode share by county

### Plot
@fig-modeshareboxplot below shows the interquartile range (IQR) of each transportation mode across all US counties. The central box represents the middle 50 percent of the data, which is the range captured by the IQR. The IQR is calculated as the difference between the third quartile (Q3) and the first quartile (Q1), the 75th and 25th percentiles of the data, respectively. The lines extending from the box, called whiskers, typically show the range of the data within 1.5 times the IQR beyond Q1 and Q3. Data points beyond the whiskers are considered outliers. When the median line sits in the middle of the box and the whiskers are similar in length, the distribution is symmetric. When the median sits closer to Q1 or Q3, and one whisker is noticeably longer than the other, the data are skewed, either positively or negatively.

```{r}
#| label: fig-modeshareboxplot
#| fig-cap: Interquartile Box Plots of Commute Mode Share among US Counties, 2024
#| eval: true
#| echo: false

modeBarButtonsList <- list("toImage")

plot_ly(modeshare_county_period_pivoted %>%
          filter(year == 2024,
                 !mode %in% c("motorcycle", "taxi", "other")) %>%
          mutate(percent = round(percent, digits = 1)),
        y = ~percent,
        color = ~reorder(mode,order),
        type = "box",
        jitter = 0.5,
        boxpoints = 'suspectedoutliers',
        quartilemethod="inclusive",
        marker = list(size = 0.75)) %>%
  config(displaylogo = FALSE,
         modeBarButtons = list(modeBarButtonsList)) %>%
  layout(yaxis = list(title = 'Commute Mode Share (%)'),
         xaxis = list(title = 'Travel Mode'),
         showlegend = FALSE)

```

### Maps
The US Census Bureau makes both attribute and geographic data available at multiple [geographic levels](https://www.census.gov/programs-surveys/economic-census/guidance-geographies/levels.html), organized according to a [standard geographic hierarchy](https://www.census.gov/programs-surveys/geography/guidance/hierarchy.html) [@census_geohierarchy]. For this exercise, we evaluate counties within their state and Census division, downloading the boundary files in R using the *tigris* package [@walker_tigris], which retrieves Census TIGER/Line shapefiles [@census_tiger]. Once downloaded, these geographies can be joined with attribute data such as commute mode share to produce meaningful maps.

TIGER/Line shapefiles are the Census Bureau's authoritative digital boundary files for states, counties, tracts and other geographies. They are used throughout the geographic information systems (GIS) field, including by transportation agencies, to map demographic, economic and infrastructure data at a consistent, nationwide scale.

@fig-modesharerates shows, county by county, the percent of workers age 16 and older who walk (a), bicycle (b), use transit (c), or drive alone (d) to work. Use the layer control in the upper right of each map to switch between the default basemap and satellite imagery, and toggle the county and division layers on or off.

```{r}
#| label: fig-modesharerates
#| fig-cap: Percent of Workers Age 16 and Older by Means of Transportation to Work by County, 2024
#| fig-subcap:
#|   - "Walk Mode Share"
#|   - "Bicycle Mode Share"
#|   - "Transit Mode Share"
#|   - "Drove Alone Mode Share"
#| eval: true
#| echo: true
#| code-fold: true

modeBarButtonsList <- list("toImage")

# plot state and county maps with leaflet
countylabels <- sprintf(
  "<strong>%s Division</strong><br/>
  <strong>%s</strong><br/>
  Total workers (16+): %s<br/>
  Walk: %0.1f%%<br/>
  Bicycle: %0.1f%%<br/>
  Transit: %0.1f%%<br/>
  Drove Alone: %0.1f%%<br/>
  Home: %0.1f%%",
  modeshare_county_latest_geom$division_name,
  modeshare_county_latest_geom$county_name,
  format(modeshare_county_latest_geom$workers16pl, big.mark=","),
  modeshare_county_latest_geom$pct_walk,
  modeshare_county_latest_geom$pct_bicycle,
  modeshare_county_latest_geom$pct_transit,
  modeshare_county_latest_geom$pct_drovealone,
  modeshare_county_latest_geom$pct_fromhome) %>%
lapply(htmltools::HTML)

# function for creating quintile bins for requested transportation mode

fx_create_bins <- function(pct_mode){
  bins_mode <- modeshare_county_latest %>%
  mutate(bins = ntile(get(pct_mode), 5)) %>%
  group_by(bins) %>%
  summarise(max = round(max(get(pct_mode)),2)) %>%
  select(max) %>%
  unlist()

  bins_mode[1] <- 0 # set initial bin to 0
  bins_mode[5] <- bins_mode[5]+0.01 # round up quintile to include all values

  return(bins_mode)
}

# function for creating mode-specific map

fx_create_map <- function(pct_mode, bincolors, legend_title) {
  # create mode-specific color palette
  pal_mode <- colorBin(bincolors, domain = modeshare_county_latest_geom[[pct_mode]], bins = fx_create_bins(pct_mode))

  # create mode-specific leaflet map
  mode_map <- leaflet(modeshare_county_latest_geom) %>%
    addTiles(urlTemplate = paste0(
      "https://basemaps.cartocdn.com/rastertiles/light_all/{z}/{x}/{y}.png?key=",
      CARTO_API_KEY
    )) %>%
    addProviderTiles(providers$Esri.WorldImagery, group = "World Imagery (satellite)") %>%
    addPolygons(
      fillColor = ~ pal_mode(get(pct_mode)),
      weight = 0.5,
      opacity = 0.75,
      color = "white",
      #dashArray = "3",
      fillOpacity = 0.75,
      label = countylabels,
      group = "Counties"
    ) %>%
    addPolylines(
      data = us_states_sub_geom,
      stroke = TRUE,
      weight = 2,
      opacity = 1,
      fillOpacity = 0,
      color = "#FFF",
      group = "States"
    ) %>%
    addPolylines(
      data = us_divisions_sub_geom,
      stroke = TRUE,
      weight = 2,
      opacity = 1,
      fillOpacity = 0,
      color = "black",
      group = "Divisions"
    ) %>%
    addLegend(
      "bottomleft",
      pal = pal_mode,
      values = ~ pct_mode,
      title = legend_title,
      labFormat = labelFormat(suffix = "%"),
      opacity = 1
    ) %>%
    addResetMapButton() %>%
    addFullscreenControl() %>%
    addLayersControl(
      baseGroups = c(# "Basemap (default)",
        "Positron (minimal)", "World Imagery (satellite)"),
      overlayGroups = c("Counties", "States", "Divisions"),
      options = layersControlOptions(collapsed = FALSE)
    )

  return(mode_map)
}

walk_map <- fx_create_map("pct_walk","Blues", "Walk Mode Share")
bicycle_map <- fx_create_map("pct_bicycle","Reds", "Bike Mode Share")
transit_map <- fx_create_map("pct_transit","Greens", "Transit Mode Share")
drove_map <- fx_create_map("pct_drovealone","Greys", "Drive Alone Mode Share")

# display maps
walk_map
bicycle_map
transit_map
drove_map

```

<!-- ::: {.callout-tip collapse="true"} -->
<!-- ## VIDEO: Step #3 -->
<!-- {{< video videos/exercise_01_03.mp4 >}} -->
<!-- ::: -->

# Step 4. Pedestrian fatalities

## Download pedestrian fatalities and examine table
Next, you will download pedestrian fatality records for your county from the Fatality Analysis Reporting System (FARS) [@nhtsa_fars], using the NHTSA CrashViewer API [@nhtsa_crashviewer]. FARS is a nationwide census, not a sample, of every crash on a US public road that results in a death within 30 days of the crash. It has been maintained by the National Highway Traffic Safety Administration (NHTSA) since 1975 and records detailed information about each crash, the vehicles involved, and the people involved, including whether a pedestrian or bicyclist was struck. Because FARS is a complete accounting of fatal crashes rather than a sample survey like the ACS, transportation planners, safety engineers and researchers rely on it to identify high injury streets and intersections, evaluate how street design relates to safety outcomes, and prioritize projects intended to reduce traffic deaths under Vision Zero and Safe System initiatives.

*Copy, paste and run the code below in your Exercise_01.R script to download pedestrian fatality data for your county using the FARS API. The code will download all fatal crash records for your selected county between 2010 and 2024, and filter those records to the crashes that involved a pedestrian.*

```{r}
#| label: download and filter FARS pedestrian fatality records
#| echo: true

state_fips   <- as.integer(substr(custom_geoid, 1, 2))
county_fips  <- as.integer(substr(custom_geoid, 3, 5))

crashes_fars <- read_csv(paste0(
  "https://crashviewer.nhtsa.dot.gov/crashviewer/CrashAPI/FARSData/GetFARSData?",
  "dataset=Accident&FromYear=2010&ToYear=2024&State=", state_fips,
  "&county=", county_fips, "&format=csv"
)) %>%
  filter(state == state_fips, county == county_fips)

ped_fatalities <- crashes_fars %>%
  filter(harm_evname == "Pedestrian") %>%
  filter(
    latitude  > -90,  latitude  < 90,
    longitud  > -180, longitud  < 180,
    !latitude  %in% c(77.7777, 88.8888, 99.9999),
    !longitud  %in% c(-77.7777, -88.8888, -99.9999)
  ) %>%
  st_as_sf(coords = c("longitudname", "latitudename"), crs = 4326)

```

## Plot of pedestrian fatalities trend
@fig-pedfatalitytrend below shows the number of pedestrian fatalities recorded in Cook County for each year between 2010 and 2024. *Copy, paste and run the code below in your Exercise_01.R to create a similar figure for your selected county. When finished, copy and paste the custom figure into slide #3 of your presentation and summarize the trend, noting any years with notable increases or decreases.*

```{r}
#| label: fig-pedfatalitytrend
#| fig-cap: Pedestrian Fatalities by Year, 2010 to 2024
#| echo: true

# Identify pedestrian-involved fatal crashes
# harm_ev == 8 (Pedestrian) captures the crash's first harmful event as a pedestrian strike;
# peds > 0 also works as a check that at least one pedestrian was involved
pedestrian_fatalities <- crashes_fars %>%
  filter(harm_ev == 8 | peds > 0) %>%
  group_by(year) %>%
  summarise(total_fatalities = sum(fatals, na.rm = TRUE),
            total_crashes = n(),
            .groups = "drop") %>%
  arrange(year)

all_fatalities <- crashes_fars %>%
  select(harm_ev, persons, year) %>% 
    mutate(collision_type = case_when(
    harm_ev == 9         ~ "Pedalcyclist",
    harm_ev == 8          ~ "Pedestrian",
    is.na(harm_ev)        ~ "Other",   # missing/unknown -- can't confirm type
    TRUE                  ~ "Motor Vehicle"
  )) %>% 
  group_by(year, collision_type) %>% 
  summarise(sum(persons)) %>% 
  arrange(year) %>% 
  ungroup()

all_fatalities <- crashes_fars %>%
  select(harm_ev, persons, year) %>%
  mutate(collision_type = case_when(
    harm_ev == 9         ~ "Pedalcyclist",
    harm_ev == 8          ~ "Pedestrian",
    is.na(harm_ev)        ~ "Other",   # missing/unknown -- can't confirm type
    TRUE                  ~ "Motor Vehicle"
  )) %>% 
  group_by(year, collision_type) %>%
  summarise(n = sum(persons, na.rm = TRUE), .groups = "drop") %>% 
  complete(year, collision_type, fill = list(n = 0)) %>%  
  arrange(year)  

colors <- c(
  "Motor Vehicle" = "#4C72B0",
  "Pedestrian"    = "#DD8452",
  "Pedalcyclist"  = "#55A868",
  "Other"         = "#C44E52"
)

# Set a consistent stacking order (bottom to top)
category_order <- c("Motor Vehicle", "Pedestrian", "Pedalcyclist", "Other")
all_fatalities$collision_type <- factor(all_fatalities$collision_type, levels = category_order)

plot_ly(
  data = all_fatalities,
  x = ~year,
  y = ~n,
  color = ~collision_type,
  colors = colors,
  type = "scatter",
  mode = "none",
  stackgroup = "one",
  hovertemplate = "%{x}<br>%{fullData.name}: %{y}<extra></extra>"
) %>%
  layout(
    title = "FARS Fatalities by Year and Collision Type",
    xaxis = list(title = "Year", dtick = 1),
    yaxis = list(title = "Number of Fatalities"),
    legend = list(title = list(text = "Collision Type")),
    hovermode = "x unified"
  )


# second plot
modeBarButtonsList <- list("toImage")

plot_ly(
  pedestrian_fatalities,
  x = ~year,
  y = ~total_fatalities,
  type = "scatter",
  mode = "lines+markers",
  line = list(color = "firebrick", width = 3),
  marker = list(color = "firebrick", size = 8)
) %>%
  layout(
    title = "Pedestrian Fatalities by Year, 2010 to 2024",
    xaxis = list(title = "Year", dtick = 1),
    yaxis = list(title = "Number of Pedestrian Fatalities"),
    hovermode = "x unified"
  ) %>% 
  config(displaylogo = FALSE,
         modeBarButtons = list(modeBarButtonsList))
```

## Maps of pedestrian fatalities over time
Before running a formal hotspot analysis, it helps to simply look at where pedestrian fatalities have occurred over time. @fig-pedfatalityperiods below locates pedestrian fatalities from your county into four multi-year periods (2010 to 2013, 2014 to 2017, 2018 to 2021 and 2022 to 2024) and maps each period side by side, with panning and zooming synced across all four panels. Copy, paste and run the following code in your custom R script. As you compare the four panels, look for areas where fatalities appear repeatedly across periods. These recurring locations are good candidates to examine further in the formal hotspot analysis that follows.

```{r}
#| label: fig-pedfatalityperiods
#| fig-cap: Pedestrian Fatalities by Four Year Period, 2010 to 2024
#| echo: true

us_county_boundary <- us_counties_population_geo %>% 
  select(NAME,
         GEOID) %>% 
  filter(GEOID == custom_geoid)

## Step A: define rolling 4 year periods across 2010 to 2024
period_breaks <- c(2010, 2014, 2018, 2022, 2025)
period_labels <- c("2010 to 2013", "2014 to 2017", "2018 to 2021", "2022 to 2024")

ped_fatalities <- ped_fatalities %>%
  mutate(period = cut(year, breaks = period_breaks, right = FALSE, labels = period_labels))

## Step B: common map extent (unname to avoid named vector JSON warnings)
county_bbox <- st_bbox(us_county_boundary)
bbox_vals <- unname(county_bbox)
names(bbox_vals) <- c("xmin", "ymin", "xmax", "ymax")

## Step C: build one map per period
period_maps <- lapply(period_labels, function(p) {

  ped_period <- ped_fatalities %>% filter(period == p)

  m <- leaflet() %>%
    addTiles(urlTemplate = paste0(
      "https://basemaps.cartocdn.com/rastertiles/light_all/{z}/{x}/{y}.png?key=",
      CARTO_API_KEY
    )) %>%
    addPolygons(
      data = us_county_boundary,
      fill = FALSE, color = "black", weight = 1.5, opacity = 1
    ) %>%
    fitBounds(
      lng1 = as.numeric(bbox_vals["xmin"]), lat1 = as.numeric(bbox_vals["ymin"]),
      lng2 = as.numeric(bbox_vals["xmax"]), lat2 = as.numeric(bbox_vals["ymax"])
    ) %>%
    addResetMapButton() %>%
    addControl(
      html = paste0("<strong>", p, "</strong> (n = ", nrow(ped_period), ")"),
      position = "topright"
    )

  if (nrow(ped_period) > 0) {
    m <- m %>%
      addCircleMarkers(
        data = ped_period, radius = 4, color = "firebrick",
        stroke = FALSE, fillOpacity = 0.7
      )
  }

  m
})

## Step D: arrange as a synced 2x2 grid (pan and zoom linked across all panels)
latticeView(
  period_maps[[1]], period_maps[[2]], period_maps[[3]], period_maps[[4]],
  ncol = 2,
  sync = "all"
)

```

## Hotspot map of pedestrian fatalities
Now run a formal hotspot analysis. The code below estimates a kernel density surface from the pedestrian fatality locations in your county, which highlights corridors and intersections with a concentration of pedestrian crashes more rigorously than eyeballing the period maps above. @fig-pedfatalityhotspot shows the relative intensity/concentration of fatalities, not raw counts such that warmer colors (orange/red) represent a larger number of spatially clustered fatalities, while lighter areas mean fewer nearby fatalities. And since the color breaks are based on natural gaps in this particular dataset, they're useful for visual comparison but shouldn't be treated as fixed, universal risk thresholds.

*Copy, paste and run the code into your Exercise #1 script. Identify any hotspots in your county. Are there areas where pedestrian fatalities appear to cluster? Copy the map into slide #4 of your presentation and describe what you find.*

```{r}
#| label: fig-pedfatalityhotspot
#| fig-cap: Pedestrian Fatality Hotspot Maps
#| fig-subcap:
#|   - "Pedestrian Fatality Density Surface"
#|   - "Pedestrian Fatality Density Surface with County Boundary"
#| echo: true

## Determine local UTM CRS from county centroid
us_county_boundary <- us_counties_population_geo %>%
  filter(GEOID == custom_geoid) %>%
  select(NAME, STATE_NAME, GEOID) %>%
  st_set_crs(4326)

county_centroid <- st_centroid(us_county_boundary)

county_centroid_zoned <- county_centroid %>%
  st_join(utm_zones %>% select(ZONE), join = st_within) %>%
  st_drop_geometry()

local_crs <- county_centroid_zoned$ZONE + 32600

## Estimate the kernel density (KDE) surface
ped_ft  <- st_transform(ped_fatalities, crs = local_crs)
ped_win <- as.owin(st_convex_hull(st_union(ped_ft)))
ped_ppp <- as.ppp(st_coordinates(ped_ft), W = ped_win)

ped_density <- density.ppp(ped_ppp, sigma = bw.diggle(ped_ppp), edge = TRUE)

# Rescale from events per square meter to events per square kilometer (x 1e6)
# so values aren't vanishingly small. This is purely for readable legend
# labels and does not change the pattern of the surface.
dens_r <- as.data.frame(ped_density) %>%
  rename(value = 3) %>%
  mutate(value = value * 1e6) %>%
  terra::rast(type = "xyz", crs = paste0("EPSG:", local_crs))

dens_r_ll <- terra::project(dens_r, "EPSG:4326", method = "bilinear")

## Clip the raster to the county boundary
county_vect <- vect(us_county_boundary)
dens_r_ll <- terra::crop(dens_r_ll, county_vect)
dens_r_ll <- terra::mask(dens_r_ll, county_vect)

## Bin density into readable classes (natural breaks on log scale)
dens_vals <- terra::values(dens_r_ll, na.rm = TRUE)

# log1p handles the near-zero values gracefully (log1p(0) = 0)
log_vals <- log1p(dens_vals)

breaks_log <- classIntervals(log_vals, n = 5, style = "jenks")$brks %>%
  unique()

# Convert breakpoints back to the original density scale for the legend/palette
breaks <- expm1(breaks_log)

dens_pal <- colorBin(
  palette  = "YlOrRd",
  domain   = dens_vals,
  bins     = breaks,
  na.color = "transparent"
)

leaflet() %>%
  addTiles(urlTemplate = paste0(
      "https://basemaps.cartocdn.com/rastertiles/light_all/{z}/{x}/{y}.png?key=",
      CARTO_API_KEY
    )) %>%
  addRasterImage(
    dens_r_ll, colors = dens_pal, opacity = 0.65,
    group = "Density Surface"
  ) %>%
  addPolygons(
    data = us_county_boundary,
    fill = FALSE, color = "darkgrey", weight = 2, opacity = 1,
    group = "County Boundary"
  ) %>%
  addCircleMarkers(
    data = ped_fatalities, radius = 3, color = "#555555", weight = 1,
    fillOpacity = 0, stroke = TRUE,
    group = "Pedestrian Fatalities"
  ) %>%
  addLegend(
    pal = dens_pal, values = dens_vals,
    title = "Pedestrian fatality<br>density (per km²)",
    position = "bottomright",
    labFormat = labelFormat(digits = 1)
  ) %>%
    addResetMapButton() %>%
    addFullscreenControl() %>%
  addLayersControl(
    overlayGroups = c("Density Surface", "County Boundary", "Pedestrian Fatalities"),
    options = layersControlOptions(collapsed = FALSE)
  )
```

::: {.callout-note collapse="false"}
## More about the hot spot map
This script computes a kernel density estimate (KDE) on the pedestrian fatality points using the *spatstat* package's [@baddeley_spatstat] density.ppp() function — essentially smoothing the individual crash locations into a continuous surface rather than leaving them as discrete dots. Each point spreads its influence outward in a soft, bell-curve-shaped bump, and where these bumps from nearby points overlap, the resulting intensity builds up, creating "hotspots" where fatalities cluster tightly together. After computing the surface, the units get converted to fatalities per km² just for readability, the raster is reprojected onto a normal map (lat/long), clipped to the county boundary, and then split into 5 color categories using the *classInt* package [@bivand_classint] to calculate Jenks natural breaks — calculated on a log-transformed version of the values first (since a few extreme hotspot cells could otherwise dominate the breaks and wash out finer patterns elsewhere).
:::

<!-- ::: {.callout-tip collapse="true"} -->
<!-- ## VIDEO: Step #4 -->
<!-- {{< video videos/exercise_01_04.mp4 >}} -->
<!-- ::: -->

# Step 5. Urban street redesign recommendations
Use the satellite view on you hot spot map and/or Google Street View to examine one of the corridors or intersections identified in @fig-pedfatalityhotspot. *What do you observe about the street design in this area? What challenges does the current design create for pedestrian safety? What infrastructure-related interventions, such as narrower travel lanes, raised crosswalks, protected bike lanes or shorter crossing distances, would you recommend for this location? Use slide #5 to summarize and illustrate your ideas.*

::: {.callout-note collapse="false"}
## Urban Street Design Resources
- [The Urban Street Design Guide](https://nacto.org/publication/urban-street-design-guide/) [@nacto_urbanstreetdesignguide], published by NACTO (National Association of City Transportation Officials), is a practitioner-focused resource that outlines design principles and tactics cities use to make streets safer, more livable, and more economically vibrant. To identify safety interventions, users can match their specific context—such as a street type (e.g., Neighborhood Street, Downtown Thoroughfare) or intersection type—to the corresponding guide section, then review targeted design elements like curb extensions, vertical speed control (speed humps/tables), crosswalk treatments, corner radii, and signal timing that address common safety issues like speeding, poor sightlines, or pedestrian conflicts. It also provides guidance on design controls (design speed, design vehicle) to calibrate interventions to context, and offers lower-cost interim strategies (temporary closures, pilot-to-permanent conversions) for testing treatments before full capital investment. Overall, the guide functions as a diagnostic-to-solution reference, letting users move from identifying a street safety problem to selecting a specific, field-tested design countermeasure.
- [The Pedestrian Program](https://www.chicago.gov/city/en/sites/complete-streets-chicago/home/pedestrian-program.html) [@cdot_pedestrianprogram] is published by the Chicago Department of Transportation (CDOT) as part of its Complete Streets Chicago initiative, aiming to build a safer, more walkable, and accessible city by prioritizing the most vulnerable street users. To identify appropriate safety interventions, users can review CDOT's data-driven approach, which targets roughly 400 intersections annually by identifying crash hot spots and developing new treatments for the most serious issues, with resources concentrated around schools, parks, and other high-traffic destinations. The page also points to supporting documents—like the Vision Zero Chicago Action Plan, the Chicago Pedestrian Plan, and a Left Turn Traffic Calming flier—that offer more specific guidance on tools, policies, and infrastructure treatments (e.g., school zone crossing improvements, traffic signal upgrades) tied to particular safety problems. Overall, the resource functions less as a technical design manual and more as a program overview, directing practitioners toward CDOT's priority areas, funding mechanisms (Capital Plan, Aldermanic menu funds), and linked plans where specific interventions and their rationale are detailed further.
:::

<!-- ::: {.callout-tip collapse="true"} -->
<!-- ## VIDEO: Step #5 -->
<!-- {{< video videos/exercise_01_05.mp4 >}} -->
<!-- ::: -->

# Step 6. Submit your completed presentation
The final step is to navigate to the course D2L page and submit your exercise presentation to the appropriate submission folder.

# References

This exercise draws on several nationwide datasets maintained by federal agencies. Each is widely used across the transportation planning, engineering and research fields, and each is worth understanding on its own terms.

**American Community Survey (ACS).** The ACS [@census_acs] is an ongoing survey administered by the US Census Bureau. It began full implementation in 2005, replacing the long form questionnaire that had previously been sent to a sample of households once every ten years as part of the decennial census. The ACS surveys roughly three and a half million addresses each year and covers social, economic, demographic and housing topics, including the "journey to work" question [@census_journeytowork] used in this exercise. Because five years of responses are pooled to produce estimates, the ACS 5 year estimates, including table B08301 (Means of Transportation to Work) [@census_b08301], are reliable even for small geographies such as counties and census tracts. Transportation planners, engineers and researchers rely on the ACS to track commuting behavior, evaluate demand for active transportation investment, and monitor progress toward sustainable transportation goals over time.

**Fatality Analysis Reporting System (FARS).** FARS [@nhtsa_fars] is a nationwide census of every crash on a US public road that results in a death within 30 days of the crash. It has been maintained by the National Highway Traffic Safety Administration (NHTSA), part of the US Department of Transportation, since 1975, and it is accessible through the NHTSA CrashViewer API [@nhtsa_crashviewer]. Because FARS records every fatal crash rather than a sample, it allows practitioners to identify high injury streets and intersections with confidence. Safety engineers, planners and researchers use FARS to evaluate how street design relates to crash outcomes and to prioritize projects intended to reduce traffic deaths, including under Vision Zero and Safe System initiatives.

**Census geographies.** The county, state and division boundaries used to map both datasets come from the Census Bureau's TIGER/Line shapefiles [@census_tiger], organized according to the Census Bureau's standard geographic hierarchy [@census_geohierarchy]. These boundaries were downloaded using the *tigris* R package [@walker_tigris], which provides direct access to Census TIGER/Line geographies from within R.

**Basemap tiles.** Every interactive map in this exercise is built on CARTO's Positron basemap tiles [@carto_basemaps], a minimal, light-colored basemap designed to keep thematic data (such as the county polygons and fatality points shown here) legible in the foreground. Each map also offers Esri's World Imagery satellite basemap [@esri_worldimagery] as an alternate layer, which is useful for examining the physical street design of the corridors and intersections identified in Step 5.

**Hotspot analysis.** The pedestrian fatality density surface in Step 4 is estimated with the *spatstat* package [@baddeley_spatstat], which implements kernel density estimation (KDE) for spatial point patterns. The resulting density values are grouped into map classes using Jenks natural breaks, calculated with the *classInt* package [@bivand_classint].

This exercise also points to several supplementary resources that are not code inputs but inform how you interpret and act on the results.

**Dangerous by Design.** *Dangerous by Design* [@sga_dangerousbydesign], published by Smart Growth America, ranks states and metro areas by pedestrian fatality rate using the same FARS and ACS data analyzed in this exercise, and is suggested as optional background reading before Step 1.

**Urban street design guidance.** The urban street redesign recommendations called for in Step 5 draw on two practitioner resources: NACTO's *Urban Street Design Guide* [@nacto_urbanstreetdesignguide], which catalogs street design treatments by context and safety issue, and the Chicago Department of Transportation's *Pedestrian Program* [@cdot_pedestrianprogram], which illustrates how one city's Complete Streets initiative applies hotspot-driven prioritization in practice.

::: {#refs}
:::
